We Are All Trying to Describe Something And We Don't Have the Words

### 2023-04-18 outputs/paper/paper.qmd
writing her up

---
title: "We Are All Trying to Describe Something And We Don't Have the Words"
subtitle: "The Jack Antonoffification of Pop Music?"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: LINK."
date: today
abstract: "In this essay, I will be exploring Jack Antonoff's production style using data from the Spotify API. Drawing from an interview with Caleb Gamman, a guy who went viral for guessing which song on Taylor Swift's Midnights album was produced by Antonoff within seconds, I will delve into the specific characteristics that define an Antonoff-y song, and assess whether or not the audio features defined by Spotify can capture this. I plan to lasso Glen McDonald,the Spotify engineer behind these features, into explaining them to me. There will be a Bayesian inference model. It will be a logistic regression. We do not know whether or not it will be good. Finally, I will armchair a little bit about the cultural significance of Antonoff's music and the reasons behind his ubiquity in the music industry."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

    > I'm literally not an expert and of course I don't actually claim to always be able to 100 identify a Jack Antonoff song it's not like I've been practicing the way that I can tell if he's produced something is basically if it sounds to me like he's produced it I know that's annoying I will clarify first I think it's mostly in the vocal treatment and that's the thing that is the most off-putting to me but just every decision uh feels a little bit wrong to me I don't know of course if I wanted to be cringe in CinemaSins I could point out the Juno 6 Bass the perfectly stacked vocals the Juno 6 pad the wide and tight plate Reverb the Juno 6 sword the goo goo gaga the man hates treble unless it's just random crap swishing around everything gets low passed with thin 707 drum samples thin 808 drum samples then 909 drum samples thin Lin drum samples thin thin drum samples all the resonances the harmonics get removed everything is sort of squished down to the fundamental reduction in Timbre everything exists in a small frequency range everything becomes a tone it's an extremely consonant sound with no interplay between the different elements now that is a skill to do and it produces a cleaner sound and the only problem is I viscerally hate that sound

@@

It is a perfect video essay. I pestered him anyway.

> "In the treble mix of the song, like in the very upper end, like sort of above where the vocals are. He often really crushes that down. Like it's lower in volume than anything else, which is sort of a weird effect. Like, when you hear someone speaking, you hear a lot of that, like, noise in their voice. When you yourself are speaking, you hear less of it, right? You hear more of your own voice bouncing around in your head. And so it sort of creates this effect of -- **like he mixes his vocals the way it would sound to the person who's singing them. Which is sort of strange.** He often has no vocals on that upper end, which is very unusual, and then often there are random little bits of noise happening up there which is the sort of thing her likes to do."

Gamman is a fan of messier music.

"Jack Antonoff does a very good job of making stuff very clean. So you know, all of the notes that are happening simulaneously are perfectly stacked. They're like mathematically stacked .. Like, you have the harmonic spectrum where everything is like a multiple on top of each other. So that's like your notes on a keyboard. And if you played a keyboard note that was a little bit off where it's supposed to be, you're gonna create this interesting effect where these two notes sort of clash against each other and they wiggle around in terms of their tone because you've got these two harmonics happening that don't match. And Jack Antonoff avoids that quite strictly"

"His drums are low past, like the vocals, so you don't get like the tinny high end in the drums or whatever because that tinny high end in the drums is like a bunch of random harmonics all over the place. So if you cut that out, you just get the tone of the drum. And then he achieves those effects using the same sounds like literally the same synthesizer on the same drum machine and stuff."

"If you used a cscope spectrogram sort of thing to look at a Jack Antonoff song, you're gonna see a lot of big lines. And then if you're listening to a song by someone else, you're gonna see like a bunch of sort of like static wiggling. It's gonna be really jagged because you've got a bunch of harmonics that don't exactly mix with each other, so they're stacking in all sorts of weird different ways"

# Numbers

@@

I pulled the data using the Spotify API, which I accessed using the `spotifyr` package in R. I used the `get_artist_audio_features` and `get_track_audio_analysis functions` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies.

@@

I chose these artists because they've produced multiple albums, and at least one album with Antonoff, as opposed to just individual songs. I chose the Chicks because I figured that collaborating with Antonoff after an extended gap in a longer spanning career than the other artists might offer some interesting results. The data points at beginning of their career reflect music that existed in a different time from the other artists in this dataset.

@@

I joined these data sets using the `rbind` function in base R, and created an additional variable called "jack", which contains a 1 if Jack Antonoff produced or co-produced the song, and a 0 if he had nothing to do with it. I got the information on which songs Antonoff did or did not produce from the "Jack Antonoff Production Discography" Wikipedia page.

@@

The `get_artist_audio_features` provides a general profile of each song, according to variables developped by spotify, which makes for simple comparisons across artists.

However, it's hard to gauge exactly what variables like "danceability" or "valence" mean, even using the documentation. For instance, "valence" is described as "A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry)". "Danceability" is described as "how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable". Neither of these definitions fully cover what metrics and thresholds go into something like "danceability". However, whatever these features capture is easier to compare across artists in visualizations than more granular details about bars, beats, duration, and tatums[^1].

[^1]: \-- a word I just learned, which refers to "the lowest regular pulse train that a listener intuitively infers from the timing of perceived musical events". The full documentation for each variable included in this study is available in the Appendix

@@

I asked Glen McDonald, a "data alchemist" (see: Principal Engineer) who worked at The Echo Nest, a machine listening start-up that was acquired by Spotify in \[YEAR\]. He still works at Spotify now.

Glen told me that Valence and Danceability we created by giving thousands of examples to college interns and asking them to tag whether a song was positive or gloomy, or danceable or un-danceable. Variables like energy or instrumentalness, on the other hand, were determined through machine listening techniques. The human subjectivity part only came in when it came to tweaking the training data. (My favourite example Glen gave was that the ML treated banjos like a human voice, giving bluegrass songs with banjos high Speechiness ratings. Because there were no banjos in the training data, they were registered as human).

Glen said that the way he's made use of these variables is by combing Energy and Valence to get a quadrant, which usually turns out "pretty good".

@@

The final audio features in this study data set contains the variables: "artist_name", "track_name", "energy", "danceability", "energy", "key", "loudness", "mode", "speechiness", "acousticness", "instrumentalness", "liveness", "valence", "tempo", "jack".

@@

# "EDA"

@@

ggplot(data = df, aes(x = energy, y = valence)) + 
  geom_point(aes(color = artist_name), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$valence), color = "gray") +
  annotate("text", x = mean(df$energy), y = max(df$valence), 
           label = "High Energy, High Valence") +
  annotate("text", x = mean(df$energy), y = min(df$valence), 
           label = "High Energy, Low Valence") +
  annotate("text", x = min(df$energy), y = mean(df$valence), 
           label = "Low Energy, High Valence") +
  annotate("text", x = max(df$energy), y = mean(df$valence), 
           label = "Low Energy, Low Valence") 
```

#| label: fig-quaddanceability
#| fig-cap: Danceability and Energy Quadrant
#| echo: false
#| warning: false

ggplot(data = df, aes(x = energy, y = danceability)) + 
  geom_point(aes(color = artist_name), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$danceability), color = "gray") +
  annotate("text", x = mean(df$energy), y = max(df$danceability), 
           label = "High Energy, High Danceability") +
  annotate("text", x = mean(df$energy), y = min(df$danceability), 
           label = "High Energy, Low Danceability") +
  annotate("text", x = min(df$energy), y = mean(df$danceability), 
           label = "Low Energy, High Danceability") +
  annotate("text", x = max(df$energy), y = mean(df$danceability), 
           label = "Low Energy, Low Danceability")

ggplot(data = df, aes(x = energy, y = danceability)) + 
  geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$danceability), color = "gray") +
  annotate("text", x = mean(df$energy), y = max(df$danceability), 
           label = "High Energy, High Danceability") +
  annotate("text", x = mean(df$energy), y = min(df$danceability), 
           label = "High Energy, Low Danceability") +
  annotate("text", x = min(df$energy), y = mean(df$danceability), 
           label = "Low Energy, High Danceability") +
  annotate("text", x = max(df$energy), y = mean(df$danceability), 
           label = "Low Energy, Low Danceability")

Sure. These are fun quadrants.But it doesn't solve the Jack-iness underlying each of those points, which are scattered pretty broadly.

@@

\[add the formula\]

This model estimates the probability P of the outcome variable Y using predictor variables Xc. In other words, it estimates the probability that a song was produced by Jack Anotnoff using the Spotfiy variables. Which gives us information about the "research" "question": are these characteristics related to Jack's music?

The chart below gives us an idea of the mean value of each of these features for Jack and non-Jack song. A logistic regression will give us a better idea of how to weigh each of these features.

@@

Following Julia Silge's tutorial \[TK\], I used the tidymodels metapackage to split the training and testing data and make cross-validation samples. The `workflow_set()` function lets us hold different sets of feature engineering together and use tuning to evaluate them all at once. I need to evaluate two elements: a basic model, and a model that was downsampled using the `themis` package.

@@

0.743 accuracy, 0.841 roc_auc. Interesting. Check it out:

@@

Danceability, with a score of 4.32, had the highest level of importance in determining whether or not a song was produced or co-produced by Jack Antonoff.

@@

Using the Spotify Audio Features, you can't really gauge what Jack Antonoff's music sound like. You get a rough idea, but you don't get the same granularity that you would by simply DM'ing Caleb Gamman on Twitter and asking him to describe it you. \[ add gamman quote\]

According to this model, though, danceability is the most important feature in predicting whether or not a song was produced by Jack Antonoff. This is so hilarious. This was a feature that was invented in a lab 15 years ago. This number refers to the time a bunch of undergraduates were paid (hopefully?) to listen to a bunch of songs and tag whether they were danceable or not. It is pretty much based on intuition. It captures this je-ne-sais-quoi essence of music that connects with listeners on a visceral level. Even if it doesn't make YOUR body want to move, you could imagine it making SOMEBODY'S body want to move.

df %>%
  filter(jack == 1) %>%
  arrange(desc(danceability)) |>
  distinct(track_name, .keep_all = TRUE) |>
  select(artist_name, danceability, track_name) |>
  knitr::kable(
     col.names = c("Artist Name", 
                   "Danceability",
                   "Track")
  )

### What do I think:

@@

    -   pretty danceable tbh

@@

    -   I don't know if "danceability" is the word, but I'll give it "anthem"-y. Like, I can't picture a group of friends dancing to this in the kitchen, but I can picture someone listening to it in their car when it comes on the radio on their way to give a presentation or negotiate a promotion or something.

@@

    -   Yeah I'm dancing. Or at least bobbing my head.

## But Danceability isn't even about Dancing! I propose: Visceral-ness

@@

One fundamental weakness is cooked into the dataset. Since we're not in the studio, we can't know the degree of Jack-iness in each individual song. Especially when he takes on the role of a more backseat co-producer, as he did with Lana Del Rey's Did You Know There's a Tunnel Under Ocean Blvd? \[tk\].

@@

@@

>>>>

---
title: "We Are All Trying to Describe Something And We Don't Have the Words"
subtitle: "The Jack Antonoffification of Pop Music?"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/jack."
date: today
abstract: "In this essay, I will be exploring Jack Antonoff's production style using data from the Spotify API. Drawing from an interview with Caleb Gamman, a guy who went viral for guessing which song on Taylor Swift's Midnights album was produced by Antonoff within seconds, I will delve into the specific characteristics that define an Antonoff-y song, and assess whether or not the audio features defined by Spotify can capture this. I plan to lasso Glen McDonald,the Spotify engineer behind these features, into explaining them to me. There will be a Bayesian inference model. It will be a logistic regression. We do not know whether or not it will be good. Finally, I will armchair a little bit about the cultural significance of Antonoff's music and the reasons behind his ubiquity in the music industry."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

    > I'm literally not an expert and of course I don't actually claim to always be able to be able to 100% identify a Jack Antonoff song. It's not like I've been practicing the way that I can tell if he's produced something. It's just basically if it sounds to me like he's produced it. I know that's annoying I will clarify first, I think it's mostly in the vocal treatment, that's the thing that is the most off-putting to me. But just every decision feels a little bit wrong to me. If I wanted to be cringe I could point out the Juno 6 Bass, the perfectly stacked vocals, the Juno 6 pad, the wide and tight plate reverb, the Juno 6 sword, the goo goo gaga, the man hates treble unless it's just random crap swishing around, everything gets low passed with thin 707 drum samples, thin 808 drum samples, then 909 drum samples, thin Lin drum samples, thin thin drum samples, all the resonances, the harmonics get removed, everything is sort of squished down to the fundamental reduction in timbre, everything exists in a small frequency range, everything becomes a tone, it's an extremely consonant sound with no interplay between the different elements. Now that is a skill to do, and it produces a cleaner sound, and the only problem is I viscerally hate that sound.

@@

It is a perfect video essay. I pestered him anyway:

> **Michaela Drouillard:** I know you cover this in the video. I'm sorry. What makes a Jack Antonoff song a Jack Antonoff song?
>
> **Caleb Gamman:** I suppose I do have much more detail that I can go into, but it is always, for me, just a knee jerk. It's not any principled thing. I do have some loose experience with like, audio production stuff. So if I really dig into it, I can maybe point some things out. In the treble mix of the song, like in the very upper end, like sort of above where the vocals are. He often really crushes that down. Like it's lower in volume than anything else, which is sort of a weird effect. Like, when you hear someone speaking, you hear a lot of that, like, noise in their voice. When you yourself are speaking, you hear less of it, right? You hear more of your own voice bouncing around in your head. And so it sort of creates this effect of -- **like he mixes his vocals the way it would sound to the person who's singing them. Which is sort of strange.** He often has no vocals on that upper end, which is very unusual, and then often there are random little bits of noise happening up there which is the sort of thing her likes to do.
>
> So I don't know, like I personally am quite a fan of messier music, you know, more experimental stuff, stuff that has a lot of clashing, sounds, a lot of different interesting pieces coming together. Jack Antonoff does a very good job of making stuff very clean. So you know, all of the notes that are happening simulaneously are perfectly stacked. They're like mathematically stacked .. Like, you have the harmonic spectrum where everything is like a multiple on top of each other. So that's like your notes on a keyboard. And if you played a keyboard note that was a little bit off where it's supposed to be, you're gonna create this interesting effect where these two notes sort of clash against each other and they wiggle around in terms of their tone because you've got these two harmonics happening that don't match. And Jack Antonoff avoids that quite strictly.
>
> His drums are low pass, like the vocals, so you don't get like the tinny high end in the drums or whatever because that tinny high end in the drums is like a bunch of random harmonics all over the place. So if you cut that out, you just get the tone of the drum. And then he achieves those effects using the same sounds like literally the same synthesizer on the same drum machine and stuff.
>
> ....
>
> If you used a cscope spectrogram sort of thing to look at a Jack Antonoff song, you're gonna see a lot of big lines. And then if you're listening to a song by someone else, you're gonna see like a bunch of sort of like static wiggling. It's gonna be really jagged because you've got a bunch of harmonics that don't exactly mix with each other, so they're stacking in all sorts of weird different ways.

I bolded the part about vocals sounding the way they would to the person who's singing them while editing this paragraph because I thought it was kind of funny, and maybe meaningful. It sounds like it's coming from inside the house!

# Numbers?

@@

I pulled the data using the Spotify API, which I accessed using the `spotifyr` package in R. I used the `get_artist_audio_features` and `get_track_audio_analysis functions` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. I threw in HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski in too, since they've never collaborated with Jack Antonoff but are more or less part of the same pop space.

@@

Of the artists who did collaborate with Antonoff, I chose these artists because they've produced multiple albums, and at least one album with Antonoff, as opposed to just individual songs. I chose the Chicks because I figured that collaborating with Antonoff after an extended gap in a longer spanning career than the other artists might offer some interesting results[^1]. The data points at beginning of their career reflect music that existed in a different time from the other artists in this dataset[^2].

[^1]: It didn't. But I kept them in there anyway because why not.

[^2]: Again, nothing worth noting came up.

@@

I joined these data sets using the `rbind` function in base R, and created an additional variable called "jack", which contains a 1 if Jack Antonoff produced or co-produced the song, and a 0 if he had nothing to do with it. I got the information on which songs Antonoff did or did not produce from the "Jack Antonoff Production Discography" Wikipedia page \[cite\].

@@

The `get_artist_audio_features` provides a general profile of each song, according to variables developped by spotify, which makes for simple comparisons across artists. The final audio features in this study data set contains the variables: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `jack`.

\[insert documentation glimpse\]

It's hard to gauge exactly what the variables "danceability" or "valence" mean, even using the documentation. For instance, "valence" is described as "A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry)". "Danceability" is described as "how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable". Neither of these definitions fully cover what metrics and thresholds go into something like "danceability". However, whatever these features capture is easier to compare across artists in visualizations than more granular details about bars, beats, duration, and tatums[^3].

[^3]: \-- a word I just learned, which refers to "the lowest regular pulse train that a listener intuitively infers from the timing of perceived musical events". The full documentation for each variable included in this study is available in the Appendix

@@

I asked Glen McDonald, a "data alchemist" (see: Principal Engineer) at Spotify who worked at The Echo Nest, a music intelligence start-up that was acquired by Spotify in 2014.

Glen told me that `valence` and `danceability` were created by giving thousands of examples to college interns and asking them to tag whether a song was positive or gloomy, or danceable or un-danceable. Variables like energy or instrumentalness, on the other hand, were determined through machine listening techniques. The human subjectivity part only came in when it came to tweaking the training data. My favourite example Glen gave was that the ML kept giving bluegrass songs insanely high `speechiness` ratings. Because there were no banjos in the training data, they were registered as human speech. (This is hilarious.) They had to go back and add more songs with banjos to the training data.

Glen said that the way he's made use of these variables is by combing energy and valence to get a quadrant, which usually turns out "pretty good".

@@

I thanked Glen for this time, ended the call, and booted up the ❤️statistical computing programming language, R ❤️\[citeR\].

ggplot(data = df, aes(x = energy, y = danceability)) + 
  geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$danceability), color = "gray") +
  theme_minimal()

ggplot(data = df, aes(x = energy, y = valence)) + 
  geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$valence), color = "gray") +
  theme_minimal()

It looks like Jack-produced songs err towards danceability, across the spectrum of energy. It's a bit harder to gauge any trends from the valence quadrant. These quadrants are a start, but they don't suggest any defining Jack-iness characteristic underlying each of those blue points, which are scattered pretty broadly.

@@

@@

@@

I made a binary response variable, called `jack`, that has a value of 1 if the song was written by Jack, and a 0 otherwise. There are 10 predictor variables, all of which are continuous: **`danceability`**, **`energy`**, **`loudness`**, **`mode`**, **`speechiness`**, **`acousticness`**, **`instrumentalness`**, **`liveness`**, **`valence`**, and **`tempo`**.

The formula for this logistic regression can be written as:

$$
P(y=1|x) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \cdots + \beta_n x_n)}}
$$

P(y=1\|x) is the probability of the dependent variable *Y* being 1, given the independent variables *x1*, *x2*, \..., *xn*, and *b0*, *b1*, *b2*, \..., *bn.* The function e\^(-(b0 + b1*x1 + b2*x2 + \... + bn\*xn)), also known as the logit function, is the exponential function of the linear combination of the independent variables and coefficients. βn represents the weights assigned to each independent variable xn. The product of βn and Xn is the contribution of the xn variable to the predicted outcome.

\[Figure X\] gives us an idea of the mean value of each of these features for Jack and non-Jack song. A logistic regression will give us a better idea of how to weigh each of these features in order to predict the outcome, which gives us information about the "research" "question": are these characteristics related to Jack's music? (And then: Which ones? And then, the most elusive question: How?)

Bracketing the major blindspot that there variables provide a single number to represent the dynamics of an entire song, the model has a few other limitations. First of all, this model is that assumes that the relationship between the independent variables and the dependent variable is linear. There could be non-linear relationships between the variables that this model isn't capturing. Second, it assumes that the relationship between the independent variables and the dependent variable is the same for all observations, which might not be true in practice -- there could be interactions between the variables, or in certain subgroups of data where the relationship is different. Albums where Jack may have had more of a back seat producing role are one example. The fact that a song's development, from inception to recording, is a whole creative process, each having its own origin story with varying degrees of Jack-ness, is another.

@@

After some minor pre-processing using the `tidyverse` and `lubridate` R packages. I followed Julia Silge's tutorial \[TK\] to build a model that deals with the imbalanced classes.

I used the `tidymodels` to fir the model. To deal with the imbalanced classes, I downsampled the majority class using the **`step_downsample()`** function from the **`themis`** package. Then I used 5-fold cross-validation with stratification by the **`jack`** variable to estimate the performance of the model.

The `workflow_set()` function lets us hold different sets of feature engineering together and use tuning to evaluate them all at once. I need to evaluate two elements: a basic model, and a model that was downsampled using the `themis` package.

I used **`rank_results()`** from the **`yardstick`** package to rank the models based on accuracy, log loss, sensitivity, and specificity. \[ask rohan about this\]

Downsampled it is.

@@

The model's accuracy score is 0.743 accuracy, and the ROC AUC score is 0.841. The accuracy score could be misleading, since the model had imbalanced classes, and downsampling may have affected the model's ability to correctly identify the minority class. The ROC AUC score is more reliable, since it's less sensitive to class imbalance.

|                            | Actually Positive (1) | Actually Negative (0) |
|----------------------------|-----------------------|-----------------------|
| **Predicted Positive (1)** | 581                   | 25                    |
| **Predicted Negative (0)** | 194                   | 51                    |

: Confusion Matrix for Logistic Regression Results

@@

Danceability, with a positive importance score of 4.32, had the highest level of importance in determining whether or not a song was produced or co-produced by Jack Antonoff. Instrumentalness comes in second at 2.59, with a negative score (meaning that negative instrumentalness is correlated), and valence comes in third with a negative score at 2.02.

df %>%
  filter(jack == 1) %>%
  arrange(desc(danceability)) |>
  distinct(track_name, .keep_all = TRUE) |>
  select(artist_name, danceability, track_name) |> 
  head(20) |>
  knitr::kable(
     col.names = c("Artist Name", 
                   "Danceability",
                   "Track")
  )

Table X shows us the 20 songs in Jack Antonoff's discography with the highest danceability scores. Bleachers takes up 35% of this list, and Taylor Swift takes up 40% of this list. St Vincent, The Chicks, Lorde, and Florence and the Machine also haves songs up there.

@@

Using the Spotify Audio Features, you can't really gauge what Jack Antonoff's music sound like. You get a rough idea, but you don't get the same granularity that you would by simply DM'ing Caleb Gamman on Twitter and asking him to describe it you. Or by watching the video where he already did that.

> I think it's mostly in the vocal treatment, that's the thing that is the most off-putting to me. But just every decision feels a little bit wrong to me. If I wanted to be cringe I could point out the Juno 6 Bass, the perfectly stacked vocals, the Juno 6 pad, the wide and tight plate reverb, the Juno 6 sword, the goo goo gaga, the man hates treble unless it's just random crap swishing around, everything gets low passed with thin 707 drum samples, thin 808 drum samples, then 909 drum samples, thin Lin drum samples, thin thin drum samples, all the resonances, the harmonics get removed, everything is sort of squished down to the fundamental reduction in timbre, everything exists in a small frequency range, everything becomes a tone, it's an extremely consonant sound with no interplay between the different elements. Now that is a skill to do, and it produces a cleaner sound, and the only problem is I viscerally hate that sound.

The Spotify Features also wouldn't help you understand in the same way you end up understanding it when you sit on my friend Matthew Jordan's couch and watch him beat box a few different Jack Antonoff and Pharrell impressions in order to show the contrast.

According to this model, though, danceability is the most important feature in predicting whether or not a song was produced by Jack Antonoff. And the accuracy metrics aren't awful -- the accuracy score is 0.743, and the roc_auc (receiver operating characteristic area under the curve) (more fun to say out loud as an abbreviation. Rock Auck), which is a metric better suited for imbalanced datasets, such as our own, is 0.810. This means that although danceability can't show us the inside of a song, something about this score is telling us something about Jack's music.

This was a feature that was invented in a lab \~15 years ago. This number refers to the time a bunch of undergraduates were paid (hopefully?) to listen to a bunch of songs and tag whether they were danceable or not. It is pretty much based on intuition. It captures this je-ne-sais-quoi essence of music that connects with listeners on a visceral level. Even if it doesn't make YOUR body want to move, you could imagine it making SOMEBODY'S body want to move.

### In which I do a close listen of our model's top 5 most danceable songs[^4]:

[^4]: I encourage you to follow along and drop me an email with your own evaluation at michaela.drouillard\@mail.utoronto.ca.

@@

    -   Pretty danceable tbh.

@@

    -   I don't know if "danceability" is the word, but I'll give it "anthem"-y. Like, I can't picture a group of friends dancing to this in the kitchen, but I can picture someone driving, on their way to give a presentation or negotiate a promotion or something, and this song comes on the radio and they leave it on. They kind of lean into it.

@@

    -   Yeah I'm dancing. Or at least bobbing my head. I only became aware of that fact when I made eye contact with the guy next to me at this cafe.

## Danceability isn't even necessarily about Dancing! I propose: Visceral-ness

@@

One fundamental weakness is cooked into the dataset. Since we're not in the studio, we can't know the degree of Jack-iness in each individual song. Especially when he takes on the role of a more backseat co-producer, as he did with Lana Del Rey's Did You Know There's a Tunnel Under Ocean Blvd? \[tk\]. We can't actually know what his involvement in each song was. There's also the issue that a study like this implies that Jack Antonoff is the reason for the way these artists' sounds develop, which takes that agency away from the artists. This is critique you see on the internet, and also a critique that some people on Twitter leveraged against Gamman after his Midnights video went viral. I mentioned it to a friend when she asked me how my final paper was going, and she said something like: "Yeah but like, it's *Taylor Swift*. As if anyone actually thinks that Jack Antonoff is the reason for her success." I agree.

@@

So another issue is that there is a lot of Taylor Swift in our training data. Taylor Swift has the second highest average danceability score across all songs, and the most danceable songs are hers. She also happens to have a lot of collaborations with Antonoff under her belt -- more than any other artist. I suspect that perhaps our model was predicting Jack via Taylor, which could be a beautiful story about collaboration and aesthetic affinity. Or it could suggest that this model has a stink to it. An avenue for future study.

@@

And if we're doing this for danceability, there might also be something worth investigating in valence, the feature that indicates a song's "positivity", which was also built via college interns' intuitions.
